---
title: "Cardiovascular Disease Risk Prediction & Machine Learning Analysis"
author: "Phyo Nyein Chan"
date: "`r Sys.Date()`"
output: html_document
---

## 1. Introduction & Project Objective
Cardiovascular diseases (CVDs) remain a leading cause of mortality globally. Identifying biological and lifestyle-related determinants is critical for early risk prediction. This study utilizes the UCI Cleveland Heart Disease dataset to identify key clinical factors influencing CVD risk, with a specific focus on **sex-specific differences** and **predictive modeling** using both traditional epidemiological statistics and advanced machine learning techniques.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(car)
library(pROC)
library(caret)
library(randomForest)
```

## 2. Data Loading & Inspection
We begin by loading the clinical dataset and inspecting its baseline characteristics and missing values to ensure statistical validity.

```{r data_loading}
# Load the dataset
heart_data <- read_csv("heart_disease_cleveland.csv")

# Check structure and count missing values
glimpse(heart_data)
cat("\nMissing Values Per Column:\n")
colSums(is.na(heart_data))
```

### *Medical Interpretation of Baseline Data:*
The dataset contains 303 patient observations. Initial data screening revealed 4 missing values in the `ca` column (number of major vessels colored by fluoroscopy) and 2 in the `thal` column. These incomplete patient records will be removed to prevent bias in subsequent statistical modeling.

```{r data_cleaning}
# Remove missing values
clean_data <- heart_data %>% drop_na(ca, thal)
```

---

## 3. Descriptive Statistics & Sex-Specific Distribution
Understanding how cardiovascular risk factors differ between male and female cohorts is central to customized risk stratification strategies.

```{r sex_visualization}
ggplot(clean_data, aes(x = as.factor(sex), fill = as.factor(target))) +
  geom_bar(position = "dodge") +
  labs(title = "Heart Disease Distribution by Sex",
       x = "Sex (0 = Female, 1 = Male)",
       y = "Patient Count",
       fill = "Heart Disease Present (1 = Yes)") +
  theme_minimal()
```

### *Medical Interpretation of the Distribution Graph:*
The bar plot reveals a distinct **sex-specific variation** in disease presentation within this sample. While the proportion of females (Sex=0) diagnosed with heart disease is relatively low, **the majority of male patients (Sex=1) exhibit active heart disease (Target=1)**. This strongly suggests that gender is an essential biological determinant in cardiovascular risk assessment.

---

## 4. Traditional Epidemiological Modeling (Logistic Regression)
To quantify the exact risk associated with clinical factors (Age, Sex, Blood Pressure, and Cholesterol), we employ Binary Logistic Regression. This method allows us to calculate **Odds Ratios (OR)**.

```{r logistic_regression}
# Fit logistic regression model
risk_model <- glm(target ~ age + sex + trestbps + chol, data = clean_data, family = binomial)
summary(risk_model)

# Compute Odds Ratios and 95% Confidence Intervals
cat("\n--- Odds Ratios (OR) and 95% CI --- \n")
odds_ratios <- exp(coef(risk_model))
conf_intervals <- exp(confint(risk_model))
cbind(Odds_Ratio = odds_ratios, conf_intervals)
```

### *Medical Interpretation of Epidemiological Model:*
The logistic regression model demonstrates substantial statistical significance for biological demographics:
- **Sex (OR ≈ 4.90, p < 0.001):** Male patients have approximately **4.9 times higher odds** of developing heart disease than female patients, even when adjusting for age, blood pressure, and cholesterol.
- **Age (OR ≈ 1.06, p < 0.001):** For every one-year increase in age, the odds of heart disease increase by roughly **6%**.
- **Trestbps & Chol:** Resting blood pressure and cholesterol showed positive risk trends but did not cross the threshold of independent statistical significance in this specific demographic model.

---

## 5. Model Diagnostics & Performance Metrics
To ensure the regression equations do not suffer from data overlap, we perform a Multi-collinearity check, alongside model evaluation using the ROC-AUC matrix.

```{r diagnostics_and_performance}
# 1. Multicollinearity Check
cat("--- Variance Inflation Factor (VIF) --- \n")
vif(risk_model)

# 2. Compute ROC-AUC Curve
clean_data$predicted_prob <- predict(risk_model, type = "response")
roc_curve <- roc(clean_data$target, clean_data$predicted_prob)

cat("\n--- AUC (Area Under Curve) --- \n")
print(auc(roc_curve))

# Plot ROC
plot(roc_curve, col = "blue", main = paste("ROC Curve (AUC =", round(auc(roc_curve), 2), ")"))
```

### *Medical Interpretation of Metrics:*
- **VIF:** All VIF values are close to 1.0, meaning there is no multicollinearity interference.
- **ROC-AUC (0.73):** An Area Under the Curve (AUC) of **0.73** indicates that this baseline model possesses a good, statistically acceptable capability to discriminate between healthy individuals and heart disease patients.

---

## 6. Advanced Machine Learning & Feature Importance
To explore non-linear interactions among clinical features, we implement a Random Forest Machine Learning classifier to automatically rank clinical variables by their diagnostic importance.

```{r machine_learning}
set.seed(123)
clean_data$target <- as.factor(clean_data$target)

# Train-Test Split (80/20)
train_index <- createDataPartition(clean_data$target, p = 0.8, list = FALSE)
train_set <- clean_data[train_index, ]
test_set  <- clean_data[-train_index, ]

# Fit Random Forest
ml_model <- randomForest(target ~ age + sex + trestbps + chol + cp + thalach + exang, data = train_set, importance = TRUE)

# View Performance
ml_predictions <- predict(ml_model, test_set)
confusionMatrix(ml_predictions, test_set$target)

# Feature Importance Plot
varImpPlot(ml_model, main = "Machine Learning: Feature Importance for Heart Disease")
```

### *Medical Interpretation of Machine Learning Findings:*
The Random Forest classifier yielded a **Balanced Accuracy of 77.37%** with a high **Sensitivity of 84.38%**. 
According to the Feature Importance plot, **Chest Pain Type (`cp`)** and **Maximum Heart Rate Achieved (`thalach`)** emerged as the two most critical clinical predictors for classification accuracy. This underscores that while non-modifiable risk factors like age and sex are highly predictive, dynamic clinical symptoms (`cp`) provide superior diagnostic classification power in active medical assessment.
